---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 150,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

In human microbiome research, there is interest in measuring “dysbiosis” which in 
many cases relies on comparing a case group with a control (without known disease). 
Several methods have been reported in literature and reviewed by Wei S et al., in their 
2021 article [Determining gut microbial dysbiosis: a review of applied indexes for assessment of intestinal microbiota imbalances](https://doi.org/10.1128/AEM.00395-21). The `dysbiosisR` provides tools for six approaches covering three different methods: Neighborhood classification, Random forest prediction and Combined alpha-beta diversity.   

It is important to note that the term 'dysbiosis' is a vaguely defined term in most instances and different methods are employed. We noticed a lack of tools for convenient and reproducible measurement of “dysbiosis” measures/scores/index. To address this gap we accumulated some commonly used approaches into a more standardized manner in an R package we call `dysbiosisR`.   
We would also like to direct the readers and users to debates surrounding the concept of dysbiosis and suggest to interpret these measures by considering its pitfalls.    

1. [Problems with the concept of gut microbiota dysbiosis](https://sfamjournals.onlinelibrary.wiley.com/doi/full/10.1111/1751-7915.13479)   

2. [Dysbiosis and Its Discontents](https://doi.org/10.1128%2FmBio.01492-17)   

3. [Dysbiosis is not an answer](https://doi.org/10.1038/nmicrobiol.2016.228)   

# Load libraries  
```{r setup}
library(dysbiosisR)
library(ggplot2)
library(microbiome)
library(dplyr)
data("WirbelJ_2018")
```

Print the available methods.  
```{r}
dysbiosisOverview()
```


Here we use the test data from [Wirbel J et al., (2018)](https://doi.org/10.1038/s41591-019-0406-6). The data consists of CRC and controls samples.  

# Data   
```{r}
WirbelJ_2018
```

```{r}
# check first five sample sums
sample_sums(WirbelJ_2018)[1:5]
```

These are relative abundance data.  

```{r}
table(sample_data(WirbelJ_2018)$disease)
```

There are 60 CRC samples and 65 control/healthy samples.  

# Dysbiosis Measures  

## Dysbiosis score based on median community level variation  
Calculates median variation in a given sample compared to a reference 
sample group. Here \code{dysbiosisMedianCLV} will calculate median 
variation for a sample compared to a reference sample set. The user 
can provide a custom distance matrix. If the user provides a Bray-Curtis 
dissimilarity matrix, then the resulting score is comparable to the 'dysbiosis score'
reported in [Lloyd-Price J, Arze C, Ananthakrishnan AN et al. (2019)](https://www.nature.com/articles/s41586-019-1237-9)

```{r}
dist.mat <- phyloseq::distance(WirbelJ_2018, "bray")
# get reference samples
ref.samples <- sample_names(subset_samples(WirbelJ_2018, 
                                           disease == "healthy"))

dysbiosis_1 <- dysbiosisMedianCLV(WirbelJ_2018,
                                  dist_mat = dist.mat,
                                  reference_samples = ref.samples)

```

The original article identifies highly divergent samples from the reference set using the 90th percentile of the dysbiosis score. This can be identified using the `quantile` function.  

```{r}
dysbiosis_thres <- quantile(subset(dysbiosis_1, disease == "CRC")$score, 0.9)
normobiosis_thres <- quantile(subset(dysbiosis_1, disease == "CRC")$score, 0.1)

dysbiosis_1 <- dysbiosis_1 |> 
  mutate(isDysbiostic = ifelse(score >= dysbiosis_thres, TRUE, FALSE))

# check in each group how many satisfy the threshold 
dysbiosis_1 |> 
  dplyr::count(disease, isDysbiostic)
```

Visualize Dysbiosis Score between CRC and healthy group.  
```{r fig.width=5, fig.height=3, fig.align='center'}
p1 <- plotDysbiosis(df=dysbiosis_1,
                    xvar="disease",
                    yvar="score",
                    colors=c(CRC="brown3", healthy="steelblue"),
                    show_points = FALSE) +
  labs(x="", y="Dysbiosis Score") +
  theme_bw(base_size = 14)
p1
```

```{r}
# order the data 
dysbiosis_1$disease <- factor(dysbiosis_1$disease, 
                              levels = c("healthy", "CRC"))

roc_1 <- pROC::roc(as.factor(dysbiosis_1$disease),
                   dysbiosis_1$score ,
                   #direction= ">",
                   plot=TRUE,
                   ci = TRUE,
                   auc.polygon=TRUE,
                   max.auc.polygon=TRUE,
                   print.auc=TRUE)
```

The AUC for this approach is 0.6469.  

## Dysbiosis score based on Euclidean distance to group centroids   

Calculates difference in `euclidean distance (ED)` for a sample 
to group centroids. For example, sample_1 to control centroid minus
sample_1 to case centroid. The user can provide a custom distance matrix. 
This approach was used in [AlShawaqfeh MK et al. (2017)](https://doi.org/10.1093/femsec/fix136).

```{r}
dysbiosis_2 <- euclideanDistCentroids(WirbelJ_2018,
                                      dist_mat = dist.mat,
                                      use_squared = TRUE,
                                      group_col = "disease",
                                      control_label = "healthy",
                                      case_label = "CRC")
dysbiosis_2[1:3,1:3]
```

Here, the `CentroidDist_score` can be used as a dysbiosis score.  

```{r}
# order the data 
dysbiosis_2$disease <- factor(dysbiosis_2$disease, 
                              levels = c("healthy", "CRC"))

roc_2 <- pROC::roc(as.factor(dysbiosis_2$disease),
                   dysbiosis_2$CentroidDist_score ,
                   #direction= ">",
                   plot=TRUE,
                   ci = TRUE,
                   auc.polygon=TRUE,
                   max.auc.polygon=TRUE,
                   print.auc=TRUE)
```

The AUC for this approach is 0.797.    

```{r fig.width=5, fig.height=3, fig.align='center'}
p3 <- plotDysbiosis(df=dysbiosis_2,
                    xvar="disease",
                    yvar="CentroidDist_score",
                    colors=c(CRC="brown3", healthy="steelblue"),
                    show_points = FALSE) +
  labs(x="", y="Dysbiosis Score (Centroid)")
p3
```

## Combined Alpha Beta Diveristy Based Score     

The combined alpha-beta diversity approach was used by [Santiago M E et al. 2019](http://www.aimspress.com/article/10.3934/microbiol.2019.1.1). This approach uses Shannon diversity as the alpha diversity measure and Jensen–Shannon divergence as the beta diversity measure. The score is mean difference of Shannon diversity between test sample and all references samples multiplied by the mean JSD of the test sample to all reference samples. When calculating this score for reference samples, the sample being used is excluded from calculating means for alpha and beta diversity.

```{r}
dysbiosis_3 <- combinedShannonJSD(WirbelJ_2018,
                                  reference_samples = ref.samples)
dysbiosis_3[1:3,1:3]
```

```{r fig.width=5, fig.height=3, fig.align='center'}
p4 <- plotDysbiosis(df=dysbiosis_3,
                    xvar="disease",
                    yvar="ShannonJSDScore",
                    colors=c(CRC="brown3", healthy="steelblue"),
                    show_points = FALSE)+
  labs(x="", y="Shannon-JSD\nDysbiosis Score")
p4
```
```{r}
# order the data 
dysbiosis_3$disease <- factor(dysbiosis_3$disease, 
                              levels = c("healthy", "CRC"))

roc_3 <- pROC::roc(as.factor(dysbiosis_3$disease),
                   dysbiosis_3$ShannonJSDScore ,
                   #direction= ">",
                   plot=TRUE,
                   ci = TRUE,
                   auc.polygon=TRUE,
                   max.auc.polygon=TRUE,
                   print.auc=TRUE)
```

The AUC for this approach is 0.582.  

## Cloud-based LOcally linear Unbiased Dysbiosis (CLOUD) test   
Cloud-based LOcally linear Unbiased Dysbiosis (CLOUD) test is a non-parametric
test and returns a measure of dysbiosis. The function was adapted from 
the original article by [Montassier E et al. 2018](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-018-0514-4). 
Here, a user defines a set of reference samples from which distance of every 
other sample is calculated. When calculating the `CLOUD` stats the _k_ is an 
important parameter specified by argument `k_num`. 

```{r}
cloud.results <- cloudStatistic(WirbelJ_2018,
                                dist_mat = dist.mat,
                                reference_samples = ref.samples,
                                ndim=-1,
                                k_num=5)
cloud.results[1:3,1:5]

```
Here, the `log2Stats` can be used as a dysbiosis score.  

```{r}
# order the data 
cloud.results$disease <- factor(cloud.results$disease, 
                                levels = c("healthy", "CRC"))

roc_4 <- pROC::roc(as.factor(cloud.results$disease),
                   cloud.results$log2Stats ,
                   #direction= ">",
                   plot=TRUE,
                   ci = TRUE,
                   auc.polygon=TRUE,
                   max.auc.polygon=TRUE,
                   print.auc=TRUE)
```

The AUC for this approach is 0.631.    

```{r fig.width=5, fig.height=3, fig.align='center'}
p2 <- plotDysbiosis(df=cloud.results,
                    xvar="disease",
                    yvar="log2Stats",
                    colors=c(CRC="brown3", healthy="steelblue"),
                    show_points = FALSE) +
  labs(x="", y="Dysbiosis CLOUD Score")
p2
```

## Random Forest Prediction Based Score  

The original article [Saffouri GB, Shields-Cutler R et al. 2019](https://doi.org/10.1038/s41467-019-09964-7) reported a Symptom Index abbreviated as SI. In this approach the feature abundances are used for Random Forest classification and the resulting out of bag (OOB) predicted probability of being classified in disease group is considered as an SI or also dysbiosis index. The `dysbiosisOBB` function in this package allows for calculating this measure with some level of freedom of `tuneRF` and random forest parameters via the `randomForest` R package.
```{r fig.width=5, fig.height=3, fig.align='center'}
# data are relative abundances summed to 100 or
# near 100 val
dysbiosis.oob <- dysbiosisOBB(WirbelJ_2018,
                              group_col = "disease",
                              case_label = "CRC",
                              seed_value = 1235,
                              add_tuneRF_params = list(ntreeTry=100,
                                                       stepFactor=1.5,
                                                       improve=0.01,
                                                       trace=TRUE,
                                                       dobest=FALSE),
                              ntree = 100,
                              plot_roc = TRUE)
```

Visualize 
```{r fig.width=5, fig.height=3, fig.align='center'}
p5 <- plotDysbiosis(df=dysbiosis.oob,
                    xvar="disease",
                    yvar="oob.score",
                    colors=c(CRC="brown3", healthy="steelblue"),
                    show_points = FALSE)+
  labs(x="", y="OOB Dysbiosis Score")
p5
```

All these measure may be influenced by host factors such as age, gender, BMI, etc. It is advised to investigate these factors when analyzing the dysbiosis measures.  

## Distance to reference plane
The original article [Halfvarson J, Brislawn CJ, Lamendella R et al. 2017](https://www.nature.com/articles/nmicrobiol20174) reported a measure based on the deviation from a healthy or reference plane. The plane is fitted in PCoA-space based on user-defined distances between samples, for example a Bray-Curtis dissimilarity matrix. Briefly, a two-dimensional plane is fitted to the first three PCoA-coordinates of samples from healthy subjects using a least-squares method. Next, the distance between a given sample to the plane is calculated, which can be considered as a measure of deviation from normal.

The original [implementation](https://github.com/ElDeveloper/reference-plane) of the reference plane calculation was written in Python by the authors of the paper. Our R implementation was benchmarked against the original code and yields identical results.

As with other measures based on a distance matrix, results can vary based on the distance chosen. Here, we applied the Bray-Curtis dissimilarity metric, whereas [Halfvarson J et al.](https://www.nature.com/articles/nmicrobiol20174) used the UniFrac distance.

```{r}
dtrp.results <- distanceToReferencePlane(
  WirbelJ_2018,
  dist_mat = dist.mat,
  reference_samples = ref.samples)
dtrp.results[1:3,1:5]
```

```{r}
# order the data 
dtrp.results$disease <- factor(dtrp.results$disease, 
                               levels = c("healthy", "CRC"))
roc_6 <- pROC::roc(as.factor(dtrp.results$disease),
                   dtrp.results$dtrpScore,
                   #direction= ">",
                   plot=TRUE,
                   ci = TRUE,
                   auc.polygon=TRUE,
                   max.auc.polygon=TRUE,
                   grid=TRUE,
                   print.auc=TRUE)
```

The AUC for this approach is 0.561.  

Visualise
```{r fig.width=5, fig.height=3, fig.align='center'}
p6 <- plotDysbiosis(df=dtrp.results,
                    xvar="disease",
                    yvar="dtrpScore",
                    colors=c(CRC="brown3", healthy="steelblue"),
                    show_points = FALSE)+
  labs(x="", y="Distance to reference plane Score")
p6
```

  
# Gradient plot  
Current dysbiosis measures do not necessarily reveal a yes and no classification of healthy and disease states. This is a challenge in gut microbiome research where individuality plays an important role. 
To demonstrate this, we can visualize this gradient with `plotDysbiosisGradient`. We use the results from the the random forest prediction based score and `dysbiosisMedianCLV` approach along with the `dysbiosis_thres` and `normobiosis_thres` values we obtained above.            
```{r fig.width=4, fig.height=5}

# define gradient cols
# browns <- c("#723d46", "#86525b", "#996770", "#ad7d86", 
#             "#c1949d", "#d6abb4", "#eac3cc", "#ede5e6")
volcano <- c("#003f5c", "#58508d","#bc5090","#ff6361", "#ffa600")

# RF OOB  
p.obb <- plotDysbiosisGradient(df=dysbiosis.oob,
                               score="oob.score",
                               high_line = 0.5,
                               # low_line = normobiosis_thres,
                               group_var = "disease",
                               group_colors=c("healthy" = "steelblue", 
                                              "CRC"= "brown3"),
                               point_size = 2,
                               bg_colors = rev(volcano),
                               jitter_width = 0.1) +
  labs(y="Dysbiosis Score", subtitle = "RF OOB Probability") +
  # adjust the x and y values to fit to plot
  ggplot2::annotate("text", x = 0.7, y = 0.55,
                    label = "OOB probability\n(0.5)", color="white")

# Median CLV    
p.clv <- plotDysbiosisGradient(df=dysbiosis_1,
                               score="score",
                               high_line = dysbiosis_thres,
                               low_line = normobiosis_thres,
                               group_var = "disease",
                               group_colors=c("healthy" = "steelblue", 
                                              "CRC"= "brown3"),
                               point_size = 2,
                               bg_colors = rev(volcano),
                               jitter_width = 0.1) +
  labs(y="Dysbiosis Score", subtitle = "Median CLV") +
  # adjust the x and y values to fit to plot
  ggplot2::annotate("text", x = 0.6, y = dysbiosis_thres+0.03,
                    label = "Dysbiosis\nThreshold", color="white")+
  ggplot2::annotate("text", x = 0.65, y = normobiosis_thres-0.03,
                    label = "Normobiosis\nThreshold", color="white")
```


```{r fig.width=8, fig.height=6}
library(patchwork)
p.obb + p.clv + plot_layout(guides = "collect")
```

Additional notes: The ROC AUC curves generated for a single data set are susceptible 
to over fitting or under fitting and therefore, the dysbiosis scores should be 
considered to be data specific. In order to build/use generalized prediction 
models based on dysbiosis measures and inferred cut-offs, different training and 
test data must be used. We see the application of these dysbiosis measure more 
specifically for single data sets where one would like to 'capture disease 
associated imbalance' into a single sample specific score. This score may be 
used to check for associations with other disease related features.    

Caution: The dysbiosis score in itself does not tell anything about the cause or 
consequence and hence careful interpretation is of paramount importance.  

Here, for example we ask how different is the gut microbiota at different CRC 
stages compared to control gut microbiota? We can look at two measures, alpha 
diversity and dysbiosis score in different CRC stages.   

```{r fig.width=7, fig.height=3}
# get alpha diversity example Shannon diversity  
sample_data(WirbelJ_2018)$shannon <- microbiome::diversity(WirbelJ_2018, "shannon")[,1]

shannon.plot <- meta(WirbelJ_2018) |> 
  mutate(crc.stage = ifelse(is.na(ajcc), "CTRL", ajcc)) |> 
  mutate(crc.stage = factor(crc.stage, levels= c("CTRL", "0", "i", "ii", "iii", "iv"))) |> 
  ggplot(aes(crc.stage, shannon)) +
  geom_boxplot(width=0.5, aes(color=disease))+
  geom_jitter(width = 0.2, aes(color=disease)) +
  theme_bw() +
  labs(x="CRC stages", y="Shannon") +
  scale_color_manual("Disease", values = c("healthy" = "steelblue", 
                                           "CRC"= "brown3"))

# random forest prediction based score  
dys.plot <- dysbiosis.oob |> 
  mutate(crc.stage = ifelse(is.na(ajcc), "CTRL", ajcc)) |> 
  mutate(crc.stage = factor(crc.stage, levels= c("CTRL", "0", "i", "ii", "iii", "iv"))) |> 
  ggplot(aes(crc.stage, oob.score)) +
  geom_boxplot(width=0.5, aes(color=disease))+
  geom_jitter(width = 0.1, aes(color=disease)) +
  theme_bw() +
  labs(x="CRC stages", y="Dysbiosis Score\n(RF OOB)") +
  scale_color_manual("Disease", values = c("healthy" = "steelblue", 
                                              "CRC"= "brown3"))
shannon.plot + dys.plot + plot_layout(guides = "collect")

```

```{r}
sessionInfo()
```



