---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  dpi = 300,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

In human microbiome research, there is interest in measuring “dysbiosis” which in 
many cases relies on comparing a case group with a control (without known disease). 
Several methods have been reported in literature and reviewed by Wei S et al., in their 
2021 article [Determining gut microbial dysbiosis: a review of applied indexes for assessment of intestinal microbiota imbalances](https://doi.org/10.1128/AEM.00395-21). The `dysbiosisR` provides tools for five approaches covering three different methods: Neighborhood classification, Random forest prediction and Combined alpha-beta diversity.   

It is important to note that the term 'dysbiosis' is a vaguely defined term in most instances and different methods are employed. We noticed a lack of tools for convenient and reproducible measurement of “dysbiosis” measures/scores/index. To address this gap we accumulated some commonly used approaches into a more standardized manner in an R package we call `dysbiosisR`.   
We would also like to direct the readers and users to debates surrounding the concept of dysbiosis and suggest to interpret these measures by considering its pitfalls.    

1. [Problems with the concept of gut microbiota dysbiosis](https://sfamjournals.onlinelibrary.wiley.com/doi/full/10.1111/1751-7915.13479)   

2. [Dysbiosis and Its Discontents](https://doi.org/10.1128%2FmBio.01492-17)   

3. [Dysbiosis is not an answer](https://doi.org/10.1038/nmicrobiol.2016.228)   

# Load libraries  
```{r setup}
library(dysbiosisR)
library(ggplot2)
library(microbiome)
library(dplyr)
data("WirbelJ_2018")
```

Here we use the test data from Wirbel J et al., (2018). The data consists of CRC and controls samples.  

# Data   
```{r}
WirbelJ_2018
```

```{r}
# check first five sample sums
sample_sums(WirbelJ_2018)[1:5]
```

These are relative abundance data.  

```{r}
table(sample_data(WirbelJ_2018)$disease)
```

There are 60 CRC samples and 65 control/healthy samples.  

# Dysbiosis Measures  

## Dysbiosis score based on median community level variation  
Calculates median variation in a given sample compared to a reference 
sample group. Here \code{dysbiosisMedianCLV} will calculate median 
variation for a sample compared to a reference sample set. The user 
can provide a custom distance matrix. If the user provides a Bray-Curtis 
dissimilarity matrix, then the resulting score is comparable to the 'dysbiosis score'
reported in [Lloyd-Price J, Arze C, Ananthakrishnan AN et al. (2019)](https://www.nature.com/articles/s41586-019-1237-9)

```{r}
dist.mat <- phyloseq::distance(WirbelJ_2018, "bray")
# get reference samples
ref.samples <- sample_names(subset_samples(WirbelJ_2018, 
                                           disease == "healthy"))

dysbiosis_1 <- dysbiosisMedianCLV(WirbelJ_2018,
                                  dist_mat = dist.mat,
                                  reference_samples = ref.samples)

```

The original article identifies highly divergent samples from the reference set using the 90th percentile of the dysbiosis score. This can be identified using the `quantile` function.  

```{r}
dysbiosis_thres <- quantile(filter(dysbiosis_1, disease == "CRC")$score, 0.9)
normobiosis_thres <- quantile(filter(dysbiosis_1, disease == "CRC")$score, 0.1)

dysbiosis_1 <- dysbiosis_1 |> 
  mutate(isDysbiostic = ifelse(score >= dysbiosis_thres, TRUE, FALSE))

# check in each group how many satisfy the threshold 
dysbiosis_1 |> 
  dplyr::count(disease, isDysbiostic)
```

Visualize Dysbiosis Score between CRC and healthy group.  
```{r fig.width=5, fig.height=3, fig.align='center'}
p1 <- plotDysbiosis(df=dysbiosis_1,
                    xvar="disease",
                    yvar="score",
                    colors=c(CRC="brown3", healthy="steelblue"),
                    show_points = FALSE) +
  labs(x="", y="Dysbiosis Score") +
  theme_bw(base_size = 14)
p1
```

```{r}
# order the data 
dysbiosis_1$disease <- factor(dysbiosis_1$disease, 
                              levels = c("healthy", "CRC"))

roc_1 <- pROC::roc(as.factor(dysbiosis_1$disease),
                   dysbiosis_1$score ,
                   #direction= ">",
                   plot=TRUE,
                   ci = TRUE,
                   auc.polygon=TRUE,
                   max.auc.polygon=TRUE,
                   grid=TRUE,
                   print.auc=TRUE)
```

The AUC for this approach is 0.6469.  

## Dysbiosis score based on Euclidean distance to group centroids   

Calculates difference in `euclidean distance (ED)` for a sample 
to group centroids. For example, sample_1 to control centroid minus
sample_1 to case centroid. The user can provide a custom distance matrix. 
This approach was used in [AlShawaqfeh MK et al. (2017)](https://doi.org/10.1093/femsec/fix136).

```{r}
dysbiosis_2 <- euclideanDistCentroids(WirbelJ_2018,
                                      dist_mat = dist.mat,
                                      use_squared = TRUE,
                                      group_col = "disease",
                                      control_label = "healthy",
                                      case_label = "CRC")
dysbiosis_2[1:3,1:3]
```

Here, the `CentroidDist_score` can be used as a dysbiosis score.  

```{r}
# order the data 
dysbiosis_2$disease <- factor(dysbiosis_2$disease, 
                                levels = c("healthy", "CRC"))

roc_2 <- pROC::roc(as.factor(dysbiosis_2$disease),
                   dysbiosis_2$CentroidDist_score ,
                   #direction= ">",
                   plot=TRUE,
                   ci = TRUE,
                   auc.polygon=TRUE,
                   max.auc.polygon=TRUE,
                   grid=TRUE,
                   print.auc=TRUE)
```

The AUC for this approach is 0.797.    

```{r fig.width=5, fig.height=3, fig.align='center'}
p3 <- plotDysbiosis(df=dysbiosis_2,
                    xvar="disease",
                    yvar="CentroidDist_score",
                    colors=c(CRC="brown3", healthy="steelblue"),
                    show_points = FALSE) +
  labs(x="", y="Dysbiosis CLOUD Score")
p3
```

## Combined Alpha Beta Diveristy Based Score     

The combined alpha-beta diversity approach was used by [Santiago M E et al. 2019](http://www.aimspress.com/article/10.3934/microbiol.2019.1.1). This approach uses Shannon diversity as the alpha diversity measure and Jensen–Shannon divergence as the beta diversity measure. The score is mean difference of Shannon diversity between test sample and all references samples multiplied by the mean JSD of the test sample to all reference samples. When calculating this score for reference samples, the sample being used is excluded from calculating means for alpha and beta diversity.

```{r}
dysbiosis_3 <- combinedShannonJSD(WirbelJ_2018,
                                  reference_samples = ref.samples)
dysbiosis_3[1:3,1:3]
```

```{r fig.width=5, fig.height=3, fig.align='center'}
p4 <- plotDysbiosis(df=dysbiosis_3,
                    xvar="disease",
                    yvar="ShannonJSDScore",
                    colors=c(CRC="brown3", healthy="steelblue"),
                    show_points = FALSE)+
  labs(x="", y="Shannon-JSD\nDysbiosis Score")
p4
```
```{r}
# order the data 
dysbiosis_3$disease <- factor(dysbiosis_3$disease, 
                                levels = c("healthy", "CRC"))

roc_3 <- pROC::roc(as.factor(dysbiosis_3$disease),
                   dysbiosis_3$ShannonJSDScore ,
                   #direction= ">",
                   plot=TRUE,
                   ci = TRUE,
                   auc.polygon=TRUE,
                   max.auc.polygon=TRUE,
                   grid=TRUE,
                   print.auc=TRUE)
```

The AUC for this approach is 0.582.  

## Cloud-based LOcally linear Unbiased Dysbiosis (CLOUD) test   
Cloud-based LOcally linear Unbiased Dysbiosis (CLOUD) test is a non-parametric
test and returns a measure of dysbiosis. The function was adapted from
the original article by [Montassier E et al. 2018](https://doi.org/10.1186/s40168-018-0514-4). 
Here, a user defines a set of reference samples from which distance of every other sample is
calculated. An important aspect to note in this function is that the method
is slightly modified to accommodate equal number of samples in reference and test
group. When calculating the \code{CLOUD} stats for reference samples the
number of neighbors is defined as number of reference samples minus 1. For
test samples, IF number of test samples matches number of samples in
reference group, then number of neighbors is also defined as number of
reference samples minus 1. This is done to avoid returning NAs in outputs
while retaining maximum number of reference samples.

```{r}

cloud.results <- cloudStatistic(WirbelJ_2018,
                                dist_mat = dist.mat,
                                reference_samples = ref.samples,
                                ndim=-1,
                                k_num=80)
cloud.results[1:3,1:5]

```
Here, the `log2Stats` can be used as a dysbiosis score.  

```{r}
# order the data 
cloud.results$disease <- factor(cloud.results$disease, 
                                levels = c("healthy", "CRC"))

roc_4 <- pROC::roc(as.factor(cloud.results$disease),
                   cloud.results$log2Stats ,
                   #direction= ">",
                   plot=TRUE,
                   ci = TRUE,
                   auc.polygon=TRUE,
                   max.auc.polygon=TRUE,
                   grid=TRUE,
                   print.auc=TRUE)
```

The AUC for this approach is 0.551.    

```{r fig.width=5, fig.height=3, fig.align='center'}
p2 <- plotDysbiosis(df=cloud.results,
                    xvar="disease",
                    yvar="log2Stats",
                    colors=c(CRC="brown3", healthy="steelblue"),
                    show_points = FALSE) +
  labs(x="", y="Dysbiosis CLOUD Score")
p2
```

## Random Forest Prediction Based Score  

The original article [Saffouri GB, Shields-Cutler R et al. 2019](https://doi.org/10.1038/s41467-019-09964-7) reported a Symptom Index abbreviated as SI. In this approach the feature abundances are used for Random Forest classification and the resulting out of bag (OOB) predicted probability of being classified in disease group is considered as an SI or also dysbiosis index. The `dysbiosisOBB` function in this package allows for calculating this measure with some level of freedom of `tuneRF` and random forest parameters via the `randomForest` R package.
```{r fig.width=5, fig.height=3, fig.align='center'}
# data are relative abundances summed to 100 or
# near 100 val
dysbiosis.oob <- dysbiosisOBB(WirbelJ_2018,
                              group_col = "disease",
                              case_label = "CRC",
                              seed_value = 1235,
                              add_tuneRF_params = list(ntreeTry=100,
                                                       stepFactor=1.5,
                                                       improve=0.01,
                                                       trace=TRUE,
                                                       dobest=FALSE),
                              ntree = 100,
                              plot_roc = TRUE)
```

Visualize 
```{r fig.width=5, fig.height=3, fig.align='center'}
p5 <- plotDysbiosis(df=dysbiosis.oob,
                    xvar="disease",
                    yvar="oob.score",
                    colors=c(CRC="brown3", healthy="steelblue"),
                    show_points = FALSE)+
  labs(x="", y="OOB Dysbiosis Score")
p5
```

Compare all five dysbiosis scores  
```{r}

all.dysbiosis <- data.frame(cbind(dysbiosis_1["score"], 
                                  dysbiosis_2["CentroidDist_score"], 
                                  dysbiosis_3["ShannonJSDScore"], 
                                  dysbiosis.oob["oob.score"], 
                                  cloud.results["log2Stats"]))
head(all.dysbiosis)
```




```{r}
sessionInfo()
```



